Bow shocks, bow waves, and dust waves

### 2017-12-24 papers/Papers1234/sec-dust-bow-intro.tex
Make a start on the cooling length section - do shock jump

We wish to calculate whether the star is capable of photoionizing the
entire bow shock shell, or whether the ionization front will be
trapped within it.  The number of hydrogen recombinations\footnote{%
  The diffuse field is treated in the on-the-spot approximation,
  assuming all emitted Lyman continuum photons are immediately
  re-absorbed locally, so the case~B recombination co-efficient,
  \(\alphaB = \num{2.6e-13}\, T_4^{-0.7}\, \si{cm^3.s^{-1}}\), is
  used, where \(T_4 = T/\SI{e4}{K}\).} %
per unit time per unit area in a fully ionized shell is
\begin{equation}
  \label{eq:shell-recombination-rate}
  \mathcal{R} = \alphaB n\shell^2 h\shell \ ,
\end{equation}
while the flux of hydrogen-ionizing photons
(\(h \nu > \SI{13.6}{eV}\)) incident on the inner edge of the shell is
\begin{equation}
  \label{eq:shell-ionizing-flux}
  \mathcal{F} = \frac{S} {4 \pi R_0^2} \ , 
\end{equation}
where \(S\) is the ionizing photon luminosity of the star.  Any shell
with \(\mathcal{R} > \mathcal{F}\) cannot be entirely photoionized by
the star, and so must have trapped the ionization front.  The column
density of the shocked shell can be found, for example, from
equations~(10) and~(12) of \citet{Wilkin:1996a} in the limit
\(v_\infty/V \to 0\) (Wilkin's parameter \(\alpha\)) and \(\theta \to 0\).  This yields
\begin{equation}
  \label{eq:shocked-shell-column}
  n\shell h\shell = \tfrac34 n R_0 \ .
\end{equation}
Assuming strong cooling behind the shock,\footnote{%
  This is shown to be justified in \S~\ref{sec:radi-cool-lengths}.
} %
the shell density is \(n\shell = \mathcal{M}_0^2 n\), where
\(\mathcal{M}_0 = v_\infty / \sound\) is the isothermal Mach number of the
external stream.\footnote{%
  \label{fn:temperature-dependence}
  The sound speed depends on the temperature and hydrogen and helium
  ionization fractions, \(y\) and \(y_{\text{He}}\) as
  \(\sound^2 = (1 + y + z_{\text{He}} y_{\text{He}}) (k T /
  \bar{m})\), where \(z_{\text{He}}\) is the helium nucleon abundance
  by number relative to hydrogen and
  \(k = \SI{1.3806503e-16}{erg.K^{-1}}\) is Boltzmann's constant.  We
  assume \(y = 1\), \(y_{\text{He}} = 0.5\), \(z_{\text{He}} = 0.09\),
  so that \(\sound = \num{11.4}\, T_4^{1/2}\, \si{km.s^{-1}}\). } %
Putting these together with equations~\eqref{eq:Rstar} and
~\eqref{eq:tau-star}, one finds that \(\mathcal{R} > \mathcal{F}\)
implies
\begin{equation}
  \label{eq:ifront-trap-x-cubed-taustar}
  x^3 \tau_* > \frac{4 S c \sound \bar{m}^2 \kappa}{3 \alpha L} \ .
\end{equation}
From equation~\eqref{eq:rad-full-x}, it can be seen that \(x\) depends
on the external stream parameters, \(n\), \(v_\infty\) only via
\(\tau_*\), and so equation~\eqref{eq:ifront-trap-x-cubed-taustar} is a
condition for \(\tau_*\), which, by using equation~\eqref{eq:taustar-typical}, becomes a condition on \(n / v_{10}^2\).  In the radiation bow shock case,
\(x = (1 + \eta)^{1/2}\), and the condition can be written:
\begin{equation}
  \label{eq:ifront-trap-density-RBS}
  \frac{n}{v_{10}^2} > \num{2.65e8} \, \frac{S_{49}^2 T_4^{3.4}}{L_4^3 (1 + \eta)^3} \ , 
\end{equation}
where
\begin{equation*}
  S_{49} = S / \bigl( \SI{e49}{s^{-1}} \bigr) \ .
\end{equation*}
Numerical values of \(S_{49}\) for our three example stars are given
in Table~\ref{tab:stars}.  In the radiation bow wave case,
\(x = 2\tau_*\), and the condition can be written:
\begin{equation}
  \label{eq:ifront-trap-taustar-bow-shock}
  \frac{n}{v_{10}^2} > \num{5.36e4} \, \frac{S_{49}^{1/2} T_4^{0.85} }{\kappa_{600}^{3/2} L_4^{3/2}} \ , 
\end{equation}

\subsection{Radiative cooling lengths}
\label{sec:radi-cool-lengths}

>>>>

In this section we calculate whether the star is capable of
photoionizing the entire bow shock shell, or whether the ionization
front will be trapped within it.  The number of hydrogen
recombinations\footnote{%
  The diffuse field is treated in the on-the-spot approximation,
  assuming all emitted Lyman continuum photons are immediately
  re-absorbed locally, so the case~B recombination co-efficient,
  \(\alphaB = \num{2.6e-13}\, T_4^{-0.7}\, \si{cm^3.s^{-1}}\), is
  used, where \(T_4 = T/\SI{e4}{K}\).} %
per unit time per unit area in a fully ionized shell is
  \label{eq:shell-recombination-rate}
  \mathcal{R} = \alphaB n\shell^2 h\shell \ ,
while the flux of hydrogen-ionizing photons
(\(h \nu > \SI{13.6}{eV}\)) incident on the inner edge of the shell is
  \label{eq:shell-ionizing-flux}
  \mathcal{F} = \frac{S} {4 \pi R_0^2} \ , 
where \(S\) is the ionizing photon luminosity of the star.  Any shell
with \(\mathcal{R} > \mathcal{F}\) cannot be entirely photoionized by
the star, and so must have trapped the ionization front.  The column
density of the shocked shell can be found, for example, from
equations~(10) and~(12) of \citet{Wilkin:1996a} in the limit
\(v_\infty/V \to 0\) (Wilkin's parameter \(\alpha\)) and \(\theta \to 0\).  This yields
  \label{eq:shocked-shell-column}
  n\shell h\shell = \tfrac34 n R_0 \ .
Assuming strong cooling behind the shock,\footnote{%
  This is shown to be justified in \S~\ref{sec:radi-cool-lengths}.
} %
the shell density is \(n\shell = \mathcal{M}_0^2 n\), where
\(\mathcal{M}_0 = v_\infty / \sound\) is the isothermal Mach number of the
external stream.\footnote{%
  \label{fn:temperature-dependence}
  The sound speed depends on the temperature and hydrogen and helium
  ionization fractions, \(y\) and \(y_{\text{He}}\) as
  \(\sound^2 = (1 + y + z_{\text{He}} y_{\text{He}}) (k T /
  \bar{m})\), where \(z_{\text{He}}\) is the helium nucleon abundance
  by number relative to hydrogen and
  \(k = \SI{1.3806503e-16}{erg.K^{-1}}\) is Boltzmann's constant.  We
  assume \(y = 1\), \(y_{\text{He}} = 0.5\), \(z_{\text{He}} = 0.09\),
  so that \(\sound = \num{11.4}\, T_4^{1/2}\, \si{km.s^{-1}}\). } %
Putting these together with equations~\eqref{eq:Rstar} and
~\eqref{eq:tau-star}, one finds that \(\mathcal{R} > \mathcal{F}\)
implies
  \label{eq:ifront-trap-x-cubed-taustar}
  x^3 \tau_* > \frac{4 S c \sound \bar{m}^2 \kappa}{3 \alpha L} \ .
From equation~\eqref{eq:rad-full-x}, it can be seen that \(x\) depends
on the external stream parameters, \(n\), \(v_\infty\) only via
\(\tau_*\), and so equation~\eqref{eq:ifront-trap-x-cubed-taustar} is a
condition for \(\tau_*\), which, by using equation~\eqref{eq:taustar-typical}, becomes a condition on \(n / v_{10}^2\).  In the radiation bow shock case,
\(x = (1 + \eta)^{1/2}\), and the condition can be written:
  \label{eq:ifront-trap-density-RBS}
  \frac{n}{v_{10}^2} > \num{2.65e8} \, \frac{S_{49}^2 T_4^{3.4}}{L_4^3 (1 + \eta)^3} \ , 
where
\begin{equation*}
  S_{49} = S / \bigl( \SI{e49}{s^{-1}} \bigr) \ .
\end{equation*}
Numerical values of \(S_{49}\) for our three example stars are given
in Table~\ref{tab:stars}.  In the radiation bow wave case,
\(x = 2\tau_*\), and the condition can be written:
  \label{eq:ifront-trap-taustar-RBW}
  \frac{n}{v_{10}^2} > \num{5.36e4} \, \frac{S_{49}^{1/2} T_4^{0.85} }{\kappa_{600}^{3/2} L_4^{3/2}} \ . 
In the wind bow shock case, the result is the same as
equation~\eqref{eq:ifront-trap-density-RBS}, but changing the factor
\((1 + \eta)^3\) to \(\eta^3\).  For the example hot stars in
Table~\ref{tab:stars}, and assuming \(\kappa_{600} = 1\),
\(T_4 = 0.8\), the resulting density threshold is
\(n > (\text{\numrange{1000}{5000}})\, v_{10}^2\), depending only
weakly on the stellar parameters, which is shown by the red lines in
Figure~\ref{fig:zones-v-n-plane}.  For the \SI{10}{M_\odot} star, this is
in the radiation bow wave regime, whereas for the higher mass stars it
is in the radiation bow shock regime.  When the external stream is
denser than this, then the outer parts of the shocked shell may be
neutral instead of ionized, giving rise to a cometary compact \hii{}
region \citep{Mac-Low:1991a, Arthur:2006a}.  This is only necessarily
true, however, when the star is isolated.  If the star is in a cluster
environment, then the contribution of other nearby massive stars to
the ionizing radiation field must be considered.

\subsection{Radiative cooling lengths}
\label{sec:radi-cool-lengths}
\newcommand\M{\ensuremath{\mathcal{M}}}
In this section, we calculate whether the radiative cooling is
sufficiently rapid behind the bow shock to allow the formation of a
thin, dense shell.  Since cooling is least efficient at low densities,
we will assume that the wind bow shock regime applies unless otherwise
specified. We label quantities just outside the shock by the subscript
``0'', quantities just inside the shock (after thermalization, but
before any radiative cooling) by the subscript ``1'', and quantities
after the gas has cooled back to the photoionization equilibrium
temperature by the subscript ``2''.  Assuming a ratio of specific
heats, \(\gamma = 5/3\), the relation between the pre-shock and immediate
post-shock quantities is
\begin{align}
  % \M_1 &= \left(\frac{\M_0^2 + 3} {5\M_0^2 - 1}\right)^{1/2} \\
  \label{eq:shock-n-jump}
  \frac{n_1}{n_0} &= \frac{4 \M_0^2} {\M_0^2 + 3} \\
  \label{eq:shock-T-jump}
  \frac{T_1}{T_0} &= \tfrac1{16} \bigl( 5\M_0^2 - 1 \bigr) \bigl( 1 + 3/\M_0^2 \bigr) \\
  \label{eq:shock-v-jump}
  \frac{v_1}{v_0} &= \left(\frac{n_1}{n_0}\right)^{-1} \ ,
\end{align}
where \(\M_0 = v_0 / \sound\).
